---
title       : The caret package
subtitle    : 
author      : Jeffrey Leek
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../librariesNew
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knitr::knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knitr::knit_hooks$set(plot = knitr:::hook_plot_html)
```

## The caret R package



[http://caret.r-forge.r-project.org/](http://caret.r-forge.r-project.org/)

---

## Caret functionality

* Some preprocessing (cleaning)
  * preProcess
* Data splitting
  * createDataPartition
  * createResample
  * createTimeSlices
* Training/testing functions
  * train
  * predict
* Model comparison
  * confusionMatrix

---

## Machine learning algorithms in R

* Linear discriminant analysis
* Regression
* Naive Bayes
* Support vector machines
* Classification and regression trees
* Random forests
* Boosting
* etc. 

---

## Why caret? 



http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf


--- 

## SPAM Example: Data splitting

```{r loadPackage}
library(caret); library(kernlab); data(spam)
inTrain <- createDataPartition(y=spam$type, p=0.75, list=FALSE) 
training <- spam[inTrain,]
testing <- spam[-inTrain,]
dim(training)
```

The createDataPartition function takes a (factor) variable and splits the data set into test and training based on p so that the sampling of different factor levels is the same in each test and training data set.

```{r}
table(training$type)
table(training$type)[[2]]/table(training$type)[[1]]
table(testing$type)
table(testing$type)[[2]]/table(testing$type)[[1]]
```


--- 

## SPAM Example: Fit a model

```{r training, dependson="loadPackage",cache=TRUE}
set.seed(32343)
modelFit <- train(type ~.,data=training, method="glm")
modelFit
```

The train function seems to work in a similar way as the fit function. Form the documentation I get this vague description: "This function sets up a grid of tuning parameters for a number of classification and regression routines, fits each model and calculates a resampling based performance measure.". So it seems to be quite a lot like fit but with additional functionaliy where you can  preproces the data and do cros-validation, etc. This is perhaps more descriptive: "train can be used to tune models by picking the complexity parameters that are associated with the optimal resampling statistics. For particular model, a grid of parameters (if any) is created and the model is trained on slightly different data for each candidate combination of tuning parameters. Across each data set, the performance of held-out samples is calculated and the mean and standard deviation is summarized for each combination. The combination with the optimal resampling statistic is chosen as the final model and the entire training set is used to fit a final model."

There are different methods that can be used. You can see all the options by typing names(getModelInfo()).

--- 

## SPAM Example: Final model

```{r finalModel, dependson="training",cache=TRUE}
modelFit <- train(type ~.,data=training, method="glm")
modelFit$finalModel
```


--- 

## SPAM Example: Prediction

This works in the same way as in the regression class. 
```{r predictions, dependson="training",cache=TRUE}
predictions <- predict(modelFit,newdata=testing)
predictions
```

--- 

## SPAM Example: Confusion Matrix

```{r confusion, dependson="predictions",cache=TRUE}
confusionMatrix(predictions,testing$type)
```

---

## Further information

* Caret tutorials:
  * [http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf](http://www.edii.uclm.es/~useR-2013/Tutorials/kuhn/user_caret_2up.pdf)
  * [http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf](http://cran.r-project.org/web/packages/caret/vignettes/caret.pdf)
* A paper introducing the caret package
  * [http://www.jstatsoft.org/v28/i05/paper](http://www.jstatsoft.org/v28/i05/paper)