---
title       : Random forests
subtitle    : 
author      : Jeffrey Leek, Assistant Professor of Biostatistics 
job         : Johns Hopkins Bloomberg School of Public Health
logo        : bloomberg_shield.png
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : tomorrow   # 
url:
  lib: ../../libraries
  assets: ../../assets
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
---


```{r setup, cache = F, echo = F, message = F, warning = F, tidy = F}
# make this an external chunk that can be included in any file
options(width = 100)
knitr::opts_chunk$set(message = F, error = F, warning = F, comment = NA, fig.align = 'center', dpi = 100, tidy = F, cache=TRUE, cache.path = '.cache/', fig.path = 'fig/')

options(xtable.type = 'html')
knitr::knit_hooks$set(inline = function(x) {
  if(is.numeric(x)) {
    round(x, getOption('digits'))
  } else {
    paste(as.character(x), collapse = ', ')
  }
})
knitr::knit_hooks$set(plot = knitr:::hook_plot_html)
```


## Random forests

1. Bootstrap samples
2. At each split, we also bootstrap the variables (that's the difference to bagging)
3. Grow multiple trees and vote

__Pros__:

1. Accuracy

__Cons__:

1. Speed
2. Interpretability
3. Overfitting


---

## Random forests

* You build a large number of trees, each based on a bootstrap sample
* At each node we allow a potenitaly difrent subset of variables to contribute to the splits
* Than we take an observation and run it through all the trees generating different predicitons
* Than we average those predications to get an aggregated probabliity


<img class=center src=assets/img/08_PredictionAndMachineLearning/forests.png height=400>

[http://www.robots.ox.ac.uk/~az/lectures/ml/lect5.pdf](http://www.robots.ox.ac.uk/~az/lectures/ml/lect5.pdf)


---

## Iris data

```{r iris, cache=TRUE}
data(iris); library(ggplot2); library(caret)
inTrain <- createDataPartition(y=iris$Species,
                              p=0.7, list=FALSE)
training <- iris[inTrain,]
testing <- iris[-inTrain,]
```


---

## Random forests

```{r forestIris, dependson="irisData",fig.height=4,fig.width=4,cache=TRUE}
modFit <- train(Species~ .,data=training,method="rf",prox=TRUE)
modFit
```

prox=TRUE generates an extra bit of information that we are going to use

---

## Getting a single tree

```{r , dependson="forestIris",fig.height=4,fig.width=4}
library(randomForest)
getTree(modFit$finalModel,k=2)
```
Here we look at a specific tree in our model. Each of the reows corresponds to a particular split. You see the left and the right doughter of the tree, based on which variable you are splitting, the value of the variable where it is split, and what the prediction is going to be out of this split.

---

## Class "centers"

```{r centers, dependson="forestIris",fig.height=4,fig.width=4}
irisP <- classCenter(training[,c(3,4)], training$Species, modFit$finalModel$prox)
irisP <- as.data.frame(irisP); irisP$Species <- rownames(irisP)
p <- qplot(Petal.Width, Petal.Length, col=Species,data=training)
p + geom_point(aes(x=Petal.Width,y=Petal.Length,col=Species),size=5,shape=4,data=irisP)
```

Here we take advantage of the prox option. This is basicaly for predicting centers of the calsses.
---

## Predicting new values

```{r predForest, dependson="centers",fig.height=4,fig.width=4,cache=TRUE}
pred <- predict(modFit,testing); testing$predRight <- pred==testing$Species
table(pred,testing$Species)
```
Look at the performance of the model.
---

## Predicting new values

```{r, dependson="predForest",fig.height=4,fig.width=4}
qplot(Petal.Width,Petal.Length,colour=predRight,data=testing,main="newdata Predictions")
```

---

## Notes and further resources

__Notes__:

* Random forests are usually one of the two top
performing algorithms along with boosting in prediction contests.
* Random forests are difficult to interpret but often very accurate. 
* Care should be taken to avoid overfitting (see [rfcv](http://cran.r-project.org/web/packages/randomForest/randomForest.pdf) funtion)


__Further resources__:

* [Random forests](http://www.stat.berkeley.edu/~breiman/RandomForests/cc_home.htm)
* [Random forest Wikipedia](http://en.wikipedia.org/wiki/Random_forest)
* [Elements of Statistical Learning](http://www-stat.stanford.edu/~tibs/ElemStatLearn/)
